Dissecting the causal question underlying the association between cancer and ADRD 

### 2021-03-23 04_reports/manuscript.Rmd
small updates

Given the previous literature on the topic, we might be interested in answering the following question: _What is the direct effect of Pin1 overexpression at age of 65 in the risk of ADRD by age 85 compared to regulated expression of Pin1 by age of 65._ To identify this effect with real-world data, we will illustrate the causal question through a directed acyclic graphs (DAGs), spelling out the data generating mechanisms and assumptions tied to the data. _Maybe say something about dags here_

Figure 1. illustrates a situation where Pin1, represented as $P$, was measured in all participants on their 65 birthday. Based on the measurement of Pin1, participant's exposure was classified as 1) Over-expressed Pin1 or 2) Regulated Pin1. Through out the follow-up each participant followed a strict monitoring for ADRD onset, represented as $Y_{85}$, and no participant was lost to follow-up nor died during the study period.

Pin1 expression (over-expressed vs. regulated) was not an intervention randomized at age 65, therefore we have to consider a set of measured covariates $L$ for which conditional exchangeability holds. Previous studies have described age, sex, educational level and race/ethnicity as the minimal adjusting set of covariates (Ospina). However, environmental and behavioral factors such as smoking, alcohol intake, or chronic diseases which are known to cause microenvironmental changes such as inflammation and changes in tissue remodelling, may translate into Pin1 over-expression and are also related to the development of ADRD. Therefore, had we measured all $L$ we could identify the direct of $P$ in $Y_{85}$ by conditioning on $L$.

@@

A major challenge related to assigning cancer diagnosis as the proxy of Pin1 is defining time zero. Not everyone who had Pin1 overexpression will be diagnosed with cancer by age 65. For this reason prospective cohort studies have considered cancer as a time-varying or time-dependent exposure. This means that a participant who had cancer diagnosis at age 70 contributes to the "regulated Pin1" from age 65 to 70, and from age 70 and onwards, this participant contributes to "Over-expressed Pin1" arm. Alternatively, studies that have included participants with cancer diagnosis at the time of the diagnosis (for example from cancer registries such as SEER) and matched participants by age. These subtleties are relevant for the next point.

Since the population is elderly, the risk of death at age 65 is not trivial and it increases over time. This means that some individuals may die prior to cancer diagnosis due to other causes, such as cardiovascular disease which is one of the leading causes of death in this age group. As such, we can only measure $P^*$ in the subset of individuals who have survived long enough to have a cancer diagnosis. We zoom in the relationship of $P \rightarrow P^*$ in Figure 3. In this figure we include 
$D = 0$ between $P$ and $P^*$, this illustrates that to observe $P^*$, we condition (box around $D = )$ on surviving long enough to have a cancer diagnosis.

>>>>

Given the previous literature on the topic, we focus the interest in answering the following question: _What is the direct effect of Pin1 over-expression at age of 65 in the risk of ADRD by age 85 compared to regulated expression of Pin1 by age of 65._ To identify this effect with real-world data, we will illustrate the causal question through a directed acyclic graphs (DAGs), spelling out the data generating mechanisms and assumptions tied to the data.

The causal graph of Figure 1 illustrates a situation where Pin1, represented as $P$, was measured in all participants on their 65 birthday. Based on the measurement of Pin1, participant's exposure was classified as 1) Over-expressed Pin1 or 2) Regulated Pin1. Through out the follow-up each participant followed a strict monitoring for ADRD onset, represented as $Y_{85}$, and no participant was lost to follow-up nor died during the study period.

was not an exposure or intervention randomized at age 65, therefore we have to consider a set of measured covariates $L$ for which conditional exchangeability holds. Previous studies have described age, sex, educational level and race/ethnicity as the minimal adjusting set of covariates (Ospina). However, environmental and behavioral factors such as smoking, alcohol intake, or chronic diseases which are known to cause microenvironmental changes such as inflammation and changes in tissue remodelling, may translate into Pin1 over-expression and are also related to the development of ADRD. Therefore, had we measured all $L$ we could identify the direct of $P$ in $Y_{85}$ by conditioning on $L$.

@@

A major challenge related to assigning cancer diagnosis as the proxy of Pin1 is defining time zero. Not everyone who had Pin1 over-expression will be diagnosed with cancer by age 65. For this reason prospective cohort studies have considered cancer as a time-varying or time-dependent exposure. This means that a participant who had cancer diagnosis at age 70 contributes to the "regulated Pin1" from age 65 to 70, and from age 70 and onwards to the "Over-expressed Pin1" arm. Alternatively, studies that have included participants with cancer diagnosis at the time of the diagnosis (for example from cancer registries such as SEER) and matched participants by age. If age at cancer diagnosis is considered as the time zero, we should be careful of adjusting for post-baseline covariates of $P$ or mediators between $P$ and $P^*$.

Since the population is elderly, the risk of death at age 65 is not trivial and it increases over time. This means that some individuals may die prior to cancer diagnosis due to other leading causes of death at this age group, such as cardiovascular disease. As such, we can only measure $P^*$ in the subset of individuals who have survived long enough to have a cancer diagnosis. We zoom in the relationship of $P \rightarrow P^*$ in Figure 3. In this figure we include 
$D = 0$ between $P$ and $P^*$, this illustrates that to observe $P^*$, we condition (box around $D = )$ on surviving long enough to have a cancer diagnosis.
